Revealing Transient Strain in Geodetic Data with Gaussian Process Regression

### 2017-06-16 HinesHetlandGPR.tex
added more

\documentclass[10pt,a4paper]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\usepackage{arydshln}
\usepackage{bm}
\usepackage{setspace}
\title{Revealing Transient Strain in Geodetic Data with Gaussian Process Regression}
\author{Trever T. Hines and Eric A. Hetland}

@@

\doublespacing

\section{Introduction}\label{sec:Introduction}
Crustal strain rates are fundamentally important quantities for assessing seismic hazard. Knowing where and how quickly strain is accumulating gives insight into where we can expect stored elastic energy to be released seismically. Consequently, secular crustal strain rates, which are estimated from GNSS data, have been used to constrain seismic hazard models such as UCERF3 \citep{Field2014}. Dense networks of continuous GNSS stations, such as the Plate Boundary Observatory (PBO), make it feasible to estimate transient crustal strain, which we consider to be any deviation from the secular strain rate.  Transient crustal strain can potentially illuminate geophysical signal that may not be immediately apparent from inspecting the GNSS displacement data. The ability to detect transient geophysical phenomena is also relevant for seismic hazard because there are instances of major earthquakes being triggered by slow slip events \citep{Roeloffs2006} and postseismic deformation \citep{Freed2001}. For these reasons, it is important to develop and improve upon methods for estimating crustal strain rates from GNSS data.

Most methods for estimating crustal strain rates from GNSS data assume some parametric form of the deformation signal. The simplest and traditional method for estimating secular crustal strain rates assumes that GNSS derived velocities can be described with a first order polynomial (i.e., having constant spatial gradients) over some subnetwork of the GNSS stations \citep[e.g][]{Frank1966,Prescott1976,Savage1986,Feigl1990,Murray2000}. The components of the strain rate tensor  for each subnetwork are then determined from the least squares fit to the observations. The assumption that deformation gradients are spatially uniform is obviously not appropriate when subnetworks span a large area.  To help overcome this deficiency, \citet{Shen1996} and \citet{Shen2015} used an inverse distance weighting scheme where the estimated strain rate at some point is primarily controlled by observations at nearby stations. However, the methods described in \citet{Shen1996} and \citet{Shen2015} are still formulated by assuming that the deformation gradients are uniform over the entire network. The errors in this assumption manifest as implausibly low formal uncertainties for the estimated strain rates. Other methods for estimating secular strain rates have parameterized GNSS derived velocities with bi-cubic splines \citep{Beavan2001}, spherical wavelets \citep{Tape2009}, and elastostatic Green's functions \citep{Sandwell2016}. The choice of parameterization can be subjective. If there are too few degrees of freedom in the parameterization then estimated strain rates will be biased and the uncertainties will be underestimated. If there are too many degrees of freedom then there will not be any coherent features in the estimated strain rates. One could parameterize deformation with a physically motivated model of interseismic deformation \citep[e.g.,][]{Meade2005,McCaffrey2007}. In such models the lithospheric rheology and plate interfaces are assumed to be known. Any errors in the assumed physical model could similarly result in biased strain estimates and underestimated formal uncertainties.

The aforementioned studies are concerned with estimating secular strain rates. In recent years the Southern California Earthquake Center (SCEC) community has shown interest in developing methods for detecting transient strain. SCEC supported a transient detection exercise \citep{Lohman2013}, where several research groups tested their methods for detecting transient geophysical signals with a synthetic GNSS dataset. Among the methods tested were the Network Strain Filter (NSF) \citep{Ohtani2010} and the Network Inversion Filter (NIF) \citep{Segall1997}. the NSF uses a wavelet parameterization to  describe the spatial component of geophysical signal and the NIF, which is intended for detecting slow fault slip, uses the elastic dislocation Green's functions from \citet{Okada1992}. For the NSF and NIF, the time dependence of the geophysical signal is modeled as integrated Brownian motion. The method described in \citet{Holt2013} was also tested in the transient detection exercise, and they calculate strain rates using a bi-cubic spatial parameterization of displacements between time epochs. \citet{Holt2013} defined a detection threshold based on the strain rate magnitude, and we will demonstrate that this is indeed an effective criterion for identifying geophysical signal. One challenge that is common to all these methods is that it is difficult to distinguish signal from noise, since the uncertainty on the inferred deformation can be significantly influenced by the choice of spatial parameterization.

In this paper we use Gaussian process regression (GPR) \citep{Rasmussen2006} to infer transient strain from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from discrete and potentially noisy data. GPR is closely related to Kriging \citep{Cressie1992} and Least Squares Collocation \citep{Kato1998}, and it has recently gained popularity as a tool for machine learning. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a normally distributed, continuous, stochastic process that is fully defined in terms of a mean function and a positive-definite covariance function. For example, Brownian motion is a well known Gaussian process in $\mathbb{R}^1$ which has zero mean and covariance function $C(t,t')=\min(t,t')$, where $t,t' \ge 0$. We use the Restricted Maximum Likelihood (REML) method to guide our choice for the prior Gaussian process.  We then incorporate the GNSS observations with the prior to form a posterior estimate of transient strain rates. The posterior transient strain rate is also a Gaussian process and, provided that our prior is sensible, has well quantified uncertainties. With well quantified uncertainties we can more confidently discern geophysical signal from noise in our estimates of transient strain rates.  We use GPR to infer transient strain rates resulting slow slip events (SSEs) in Cascadia. Our results demonstrate that GPR can be an effective tool for detecting SSEs and revealing the spatio-temporal evolution of crustal strain during SSEs.

@@

We consider three potential models for the temporal covariance of $u$. First, we consider the one-dimensional SE covariance function, 
\begin{equation}\label{eq:TimeSE}
T(t,t') = \phi^2\exp\left(\frac{-|t - t'|^2}{\theta^2}\right),
\end{equation}
because the SE covariance function is versatile and commonly used for non-parametric modeling. Note that $T$ includes the hyperparameter $\phi$, which serves to scale the covariance function $C_u$. Second, we consider integrated Brownian motion (IBM). IBM has zero mean and its covariance function can be found by integrating the covariance function for Brownian motion as
\begin{align}\label{eq:IBM}
T(t,t') &= \int_0^t \int_0^{t'} \phi^2 \min(\tau,\tau') \,d\tau'\,d\tau \\
        &= \frac{\phi^2}{2}\min(t,t')^2 \left(\max(t,t') - \frac{1}{3}\min(t,t')\right), \ \ \ t,t' \geq 0.
\end{align}
IBM has been used, in the context of Kalman filtering, as a non-parametric model for the time dependence of geophysical signals \citep[e.g.,][]{Segall1997,McGuire2003,Ohtani2010,Hines2016}. It should be emphasized $t=0$ is a reference time at which the Gaussian process is exactly zero.  For some geophysical signals, it is appropriate to have this reference time. For example, if we are trying to identify postseismic deformation then $t$ should be zero at the time of the earthquake.  However, if we are interesting in detecting transient events, where there is no known start time, then IBM may not be an appropriate prior and an isotropic Gaussian process should be preferred. In the following analysis, we make the quite arbitrary choice that $t$ is zero on the first epoch of $\bm{d}_*$.  Our third option for $T$ is the Wendland class of covariance functions \citep{Wendland2005}. Wendland covariance functions have compact support and hence their corresponding covariance matrices will be sparse. In our analysis, we exploit this sparsity with the CHOLMOD software package \citep{Chen2008}.  Wendland functions are positive definite in $\mathbb{R}^d$, and they describes an isotropic Gaussian process with realizations that can be differentiated $k$ times. The form of the covariance function depends on the choice of $d$ and $k$. To describe the temporal covariance of $u$, we only require that $d=1$ and $k=1$. The corresponding Wendland covariance function is 
\begin{equation}\label{eq:Wendland}
T(t,t') = \phi^2\left(1 - \frac{|t - t'|}{\theta}\right)^3_+ \left(\frac{3|t - t'|}{\theta} + 1\right), 
\end{equation}
where
\begin{equation}
(t)_+ = 
\begin{cases}
t, \ \ \ t > 0 \\
0, \ \ \ \mathrm{otherwise}.
\end{cases}
\end{equation}

@@

>>>>

\documentclass[10pt,letter]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\usepackage{arydshln}
\usepackage{bm}
\usepackage{setspace}
\title{Revealing Transient Strain in Geodetic Data with Gaussian Process Regression}
\author{Trever T. Hines and Eric A. Hetland}

@@

%\doublespacing

\section{Introduction}\label{sec:Introduction}
Crustal strain rates are fundamentally important quantities for assessing seismic hazard. Knowing where and how quickly strain is accumulating gives insight into where we can expect stored elastic energy to be released seismically. Consequently, secular crustal strain rates, which can be estimated from GNSS data, have been used to constrain seismic hazard models such as UCERF3 \citep{Field2014}. Dense networks of continuous GNSS stations, such as the Plate Boundary Observatory (PBO), make it feasible to estimate transient crustal strain, which we consider to be any deviation from the secular strain rate.  Transient crustal strain can potentially illuminate geophysical signal that may not be immediately apparent from inspecting the GNSS displacement data. The ability to detect transient geophysical phenomena is also relevant for seismic hazard because there are instances of major earthquakes being triggered by slow slip events \citep{Roeloffs2006} and postseismic deformation \citep{Freed2001}. For these reasons, it is important to develop and improve upon methods for estimating crustal strain rates from GNSS data.

Most methods for estimating crustal strain rates from GNSS data assume some parametric form of the deformation signal. The simplest and traditional method for estimating secular crustal strain rates assumes that GNSS derived velocities can be described with a first order polynomial (i.e., having constant spatial gradients) over some subnetwork of the GNSS stations \citep[e.g][]{Frank1966,Prescott1976,Savage1986,Feigl1990,Murray2000}. The components of the strain rate tensor  for each subnetwork are then determined from the least squares fit to the observations. The assumption that deformation gradients are spatially uniform is obviously not appropriate when subnetworks span a large area.  To help overcome this deficiency, \citet{Shen1996} and \citet{Shen2015} used an inverse distance weighting scheme where the estimated strain rate at some point is primarily controlled by observations at nearby stations. However, the methods described in \citet{Shen1996} and \citet{Shen2015} are still formulated by assuming that the deformation gradients are uniform over the entire network. The errors in this assumption manifest as implausibly low formal uncertainties for the estimated strain rates. Other methods for estimating secular strain rates have parameterized GNSS derived velocities with bi-cubic splines \citep{Beavan2001}, spherical wavelets \citep{Tape2009}, and elastostatic Green's functions \citep{Sandwell2016}. The type of basis functions and the number of degrees of freedom for a parameterization can be subjective. If there are too few degrees of freedom in the parameterization then estimated strain rates will be biased and the uncertainties will be underestimated. If there are too many degrees of freedom then there will not be any coherent features in the estimated strain rates. The methods described by \citet{Beavan2001}, \citet{Tape2009}, and \citet{Sandwell2016} also require the user to specify penalty parameters that control a similar trade-off between bias and variance in the solution. One could parameterize deformation with a physically motivated model of interseismic deformation \citep[e.g.,][]{Meade2005,McCaffrey2007}. In such models the lithospheric rheology and plate interfaces are assumed to be known. Any errors in the assumed physical model could similarly result in biased strain estimates and underestimated formal uncertainties.

The aforementioned studies are concerned with estimating secular strain rates. In recent years the Southern California Earthquake Center (SCEC) community has shown interest in developing methods for detecting transient strain. SCEC supported a transient detection exercise \citep{Lohman2013}, where several research groups tested their methods for detecting transient geophysical signals with a synthetic GNSS dataset. Among the methods tested were the Network Strain Filter (NSF) \citep{Ohtani2010} and the Network Inversion Filter (NIF) \citep{Segall1997}. the NSF uses a wavelet parameterization to  describe the spatial component of geophysical signal and the NIF, which is intended for detecting slow fault slip, uses the elastic dislocation Green's functions from \citet{Okada1992}. For the NSF and NIF, the time dependence of the geophysical signal is modeled as integrated Brownian motion. The method described in \citet{Holt2013} was also tested in the transient detection exercise, and they calculate strain rates using a bi-cubic spatial parameterization of displacements between time epochs. \citet{Holt2013} defined a detection threshold based on the strain rate magnitude, and we will demonstrate that this is indeed an effective criterion for identifying geophysical signal. For the same reasons descibed above, the transient deformation and corresponding uncertainties estimated by these methods can be biased by the chosen spatial parameterization. It is then difficult to distinguish signal from noise with these methods, which limits their utility for transient detection.

In this paper we use Gaussian process regression (GPR) \citep{Rasmussen2006} to infer transient strain from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered and potentially noisy data. GPR is ideal for synthesizing GNSS data into a spatially and temporally continuous represention of surface deformation, since GNSS stations are irregularly spaced and observation times may differ between stations. GPR is closely related to Kriging \citep{Cressie1992} and Least Squares Collocation \citep{Kato1998}, and it has recently gained popularity as a tool for machine learning. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a normally distributed, continuous, stochastic process that is fully defined in terms of a mean function and a positive-definite covariance function. For example, Brownian motion is a well known Gaussian process in $\mathbb{R}^1$ which has zero mean and covariance function $\mathrm{cov}(t,t')=\min(t,t')$, where $t,t' \ge 0$. If no prior information is available for the geophysical signal, then maximum likelihood methods can be used to choose a prior Gaussian process which is most consistent with the observations.  We incorporate the GNSS observations with the prior to form a posterior estimate of transient strain. The posterior transient strain is also a Gaussian process and, provided that our prior is sensible, has well quantified uncertainties. With well quantified uncertainties we can more confidently discern geophysical signal from noise in our estimates of transient strain.  We use GPR to infer transient strain resulting from slow slip events (SSEs) in Cascadia. Our results demonstrate that GPR can be an effective tool for detecting SSEs and revealing the spatio-temporal evolution of crustal strain during SSEs.

@@

We consider three potential models for the temporal covariance of $u$. First, we consider the one-dimensional SE covariance function, 
\begin{equation}\label{eq:TimeSE}
T(t,t') = \phi^2\exp\left(\frac{-|t - t'|^2}{\theta^2}\right),
\end{equation}
because the SE covariance function is versatile and commonly used for non-parametric modeling. Note that $T$ includes the hyperparameter $\phi$, which serves to scale the covariance function $C_u$. Second, we consider integrated Brownian motion (IBM). IBM has zero mean and its covariance function can be found by integrating the covariance function for Brownian motion as
\begin{align}\label{eq:IBM}
T(t,t') &= \int_0^t \int_0^{t'} \phi^2 \min(\tau,\tau') \,d\tau'\,d\tau \\
        &= \frac{\phi^2}{2}\min(t,t')^2 \left(\max(t,t') - \frac{1}{3}\min(t,t')\right), \ \ \ t,t' \geq 0.
\end{align}
IBM has been used, in the context of Kalman filtering, as a non-parametric model for the time dependence of geophysical signals \citep[e.g.,][]{Segall1997,McGuire2003,Ohtani2010,Hines2016}. It should be emphasized $t=0$ is a reference time at which the Gaussian process is exactly zero.  For some geophysical signals, it is appropriate to have this reference time. For example, if we are trying to identify postseismic deformation then $t$ should be zero at the time of the earthquake.  However, if we are interesting in detecting transient events, where there is no known start time, then IBM may not be an appropriate prior and an isotropic Gaussian process should be preferred. In the following analysis, we make the quite arbitrary choice that $t$ is zero on the first epoch of $\bm{d}_*$. Using an earlier reference time does not change the results discussed in this section. Our third option for $T$ is the Wendland class of covariance functions \citep{Wendland2005}. Wendland covariance functions have compact support and hence their corresponding covariance matrices will be sparse. In our analysis, we exploit this sparsity with the CHOLMOD software package \citep{Chen2008}.  Wendland functions are positive definite in $\mathbb{R}^d$, and they describes an isotropic Gaussian process with realizations that can be differentiated $k$ times. The form of the covariance function depends on the choice of $d$ and $k$. To describe the temporal covariance of $u$, we only require that $d=1$ and $k=1$. The corresponding Wendland covariance function is 
\begin{equation}\label{eq:Wendland}
T(t,t') = \phi^2\left(1 - \frac{|t - t'|}{\theta}\right)^3_+ \left(\frac{3|t - t'|}{\theta} + 1\right), 
\end{equation}
where
\begin{equation}
(t)_+ = 
\begin{cases}
t, \ \ \ t > 0 \\
0, \ \ \ \mathrm{otherwise}.
\end{cases}
\end{equation}

@@

% We could have also used a spherical spatial covariance functions
